Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-30 doc/thesis/methods_results.Rnw
A bit more writing

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
    \State $T \gets$ a single node with label $1$
    \State $\tip[v] \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, \beta + \gamma) < \beta$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
          \State $\tip[v] \gets |T|-1$
          \State $\tip[u] \gets |T|$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State delete $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees, via the
dot-product of the two trees' feature vectors in the infinite-dimensional space
of all possible subset trees with branch lengths (see \cref{subsec:treeshape}).
The kernel was implemented using the fast algorithm developed by
\textcite{moschitti2006making}. First, the production rule of each node, which
is the total number of children and the number of leaf children, is recorded.
The nodes of both trees are ordered by production rule, and a list of pairs of
nodes sharing the same production rule is created. These are the nodes for
which the value of the tree kernel must be computed - all other pairs have a
value of zero. The pairs to be compared are then re-ordered so that the child
nodes are always evaluated before their parents. Due to its recursive
definition, ordering the pairs in this way allows the tree kernel to be
computed by dynamic programming. The complexity of this implementation is
$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.

The tree kernel cannot be used directly as a distance measure for \gls{ABC},
since it is maximized, not minimized, when the two trees being compared are the
same. Therefore, we defined the distance between two trees as
\[
  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
\]
which is a number between 0 and 1 minimized when $T_1 = T_2$. This is similar
to the normalization used by \textcite{collins2002new, poon2013mapping}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, and is mitigated by the
adaptive approach. A single parameter $\alpha$ (not to be confused with the
\gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$ being
chosen to satisfy
\[
  \ESS(w_i) = \alpha \ESS(w_{i-1}).
\]
Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets which are still inside the new threshold. The algorithm may
be stopped when one of two termination conditions is reached. The user may
specify a final tolerance $\varepsilon$, or a final acceptance rate of the
\gls{MCMC} kernel. The latter condition stops the algorithm when the particles
are not moving around very much, implying little change in the estimated
target.

@@

We assumed that all contacts had symmetric transmission risk, which was
implemented by replacing each undirected edge in the network with two directed
edges (one in each direction). Nodes in our networks followed simple \gls{SI}
dynamics, meaning that they became infected at a rate proportional to their
number of infected neighbours, and never recovered. We did not consider the
time scale of the transmission trees in these simulations, only their shape.
Therefore, the transmission rate along each edge in the network was set to 1,
the removal rate of each node was set to 0, and all transmission trees' branch
lengths were scaled by their mean. \textit{igraph} library's implementation of
the BA model~\autocite{csardi2006igraph} was used to generate the graphs. The
analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
computer cluster. With the exception of our own programs, all analyses were
done in \software{R}, and all packages listed below are \software{R} packages.
{\color{blue}\uline{Code to run all simulation-based experiments is freely
available at \url{https://github.com/rmcclosk/thesis}.}}

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters might be identifiable from the shape of the
transmission tree. TODO }

\uline{In addition, a secondary objective of this section was to validate the
use of the tree kernel as a distance measure for \gls{ABC} in our context. As
discussed in} \cref{sec:abc}, \uline{the choice of distance function is
extremely important for the accuracy of the \gls{ABC} approximation to the
posterior. Therefore, we evaluated two additional tree statistics in the same
manner as we evaluated the tree kernel (that is, by constructing and testing a
classifier). First, we considered Sackin's index~\autocite{shao1990tree}, which
measures the degree of imbalance or asymmetry in a phylogeny} (see
\cref{subsec:treeshape}). \uline{Sackin's index is widely used for
characterizing phylogenies~\autocite{frost2013modelling} and has been
demonstrated to vary between transmission trees simulated under different
contact network types~\autocite{leventhal2012inferring}. Sackin's index does
not take branch lengths into account, considering only the tree's topology. The
other statistic we considered was the
\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
normalized distributions of their branching times} (see \cref{sec:treeshape}).
\uline{In contrast with Sackin's index, the \gls{nltt} does not explicity
consider the trees' topologies, but it does use their normalized branch
lengths. While the \gls{nltt} is a newly developed statistic not yet in
widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
basis of seminal early work extracting epidemiological information from
phylodynamics~\autocite{holmes1995revealing}.}

\uline{We expect the tree kernel to classify the \gls{BA} parameters more
accurately than either Sackin's index or the \gls{nltt}, since the tree kernel
takes both topology and branch lengths into account. }}

This experiment involved a large number of variables which were varied
combinatorially. For ease of exposition, we will describe a single experiment
first, then enumerate the values of all variables for which the experiment was
repeated. The parameters of the tree kernel, $\lambda$ and $\sigma$
(\cref{subsec:treeshape}) will be referred to as \defn{meta-parameters} to
distinguish them from the parameters of the \gls{BA} model.

The attachment power parameter \gls{alpha} was varied among three values: 0.5,
1.0, and 1.5. For each value, the \software{sample\_pa} function in the
\software{igraph} package was used to simulate 100 networks, with the other
parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
300 networks. An epidemic was simulated on each network using our
\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
point 500 of them were sampled to form a transmission tree. A total of 300
transmission trees were thus obtained, comprised of 100 trees for each of the
three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
descending from the left child of each node was not smaller than that
descending from the right child. Summary statistics, such as Sackin's index and
the ratio of internal to terminal branch lengths, were computed for each
simulated tree using our \software{treestat} binary. The trees were visualized
using the \software{ape} package~\autocite{paradis2004ape}. Our
\software{treekernel} binary was used to calculate the value of the kernel for
each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and $\sigma
= 4$. These values were stored in a symmetric 300 $\times$ 300 kernel matrix.
Similarly, we computed the \gls{nltt} statistic between each pair of trees
using our \software{treestat} binary, and stored them in a second $300 \times
300$ matrix.

@@

We simulated three trees each under a variety of parameter values, and ran the
\software{netabc} program to estimate posterior distributions for the
parameters. The parameter values and priors used are listed in
\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\lambda = 0.3$
and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
sampled datasets per particle, and the $\alpha$ parameter (not to be confused
with the \gls{BA} preferential attachment parameter, see
\cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
acceptance rate of the \gls{MH} kernel dropped below 1.5\%, the same criterion
used by \citeauthor{del2012adaptive}. Approximate marginal posterior densities
for each parameter were calculated using the \software{density} function in
\software{R} applied to the final weighted population of particles.
{\color{red}\sout{Credible intervals were obtained for each parameter using the
\software{HPDinterval} function in the \software{coda}
package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
obtained for each parameter using the \software{wtd.mean} function in the
\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
obtained using the \software{hpd} function in the \software{TeachingDemos}
package~\autocite{snow2013teachingdemos}.}}

@@

Two further experiments were performed to address potential sources of error.
To evaluate the effect of model misspecification in the case of heterogeneity
among nodes, we generated a network where half the nodes were attached with
power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
investigate the effects of potential sampling
bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
the tips were sampled in a peer-driven fashion, rather than at random. That is,
the probability to sample a node was twice as high if any of that node's
network peers had already been sampled. The parameters of this network were $N$
= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.

@@

Trees simulated under different values of \gls{alpha} were visibly quite
distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes which, once
infected, are likely to transmit to many other nodes. This results in a more
unbalanced, ladder-like structure in the phylogeny, compared to networks with
lower \gls{alpha} values. None of the other three parameters produced trees
which were as easily distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).  Sackin's index, which
measures tree imbalance, was significantly correlated with all four parameters
    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
     \Sexpr{round(I.sackin.cor$estimate, 2)},
     \Sexpr{round(m.sackin.cor$estimate, 2)},
     \Sexpr{round(N.sackin.cor$estimate, 2)};
     $p$-values
     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
     $\Sexpr{pp(I.sackin.cor$p.value)}$,
     $\Sexpr{pp(m.sackin.cor$p.value)}$,
     $\Sexpr{pp(N.sackin.cor$p.value)}$).
The ratio of internal to terminal branch lengths was negatively correlated with
\gls{alpha} and \gls{I}, and positively correlated with \gls{m} and \gls{N}
  (Spearman's rho
    \Sexpr{round(alpha.ratio.cor$estimate, 2)},
    \Sexpr{round(I.ratio.cor$estimate, 2)},
    \Sexpr{round(m.ratio.cor$estimate, 2)},
    \Sexpr{round(N.ratio.cor$estimate, 2)};
  all $p < 10^{-5}$).

@@

\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
first two principal components of the kernel matrix. The figure shows only the
simulations with 500-tip trees and 1000 infected nodes. The three \gls{alpha}
and \gls{I} values considered are well separated from each other in feature
space. On the other hand, the three \gls{N} values overlap significantly, and
the three \gls{m} values are virtually indistinguishable. Similar observations
can be made for other values of \gls{I} and the number of tips
(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
\gls{N} separated more clearly with larger numbers of tips, and in the case of
\gls{N}, larger epidemic sizes.

@@

The accuracy of classifiers $I$ varied significantly with the number of tips in
the tree. For 100-tip trees, the average $R^2$ values were
  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
trees, the values increased to
  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
Finally, the performance of classifiers for $N$ depended heavily on the
epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
for the smallest epidemic and smallest sample size, to
  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
to
  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
Sackin's index did not accurately classify $N$ in any scenario, with an average
$R^2$ of
  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
and little variation between scenarios.

>>>>

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
    \State $T \gets$ a single node with label $1$
    \State $\tip[v] \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, \beta + \gamma) < \beta$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State $\tip[v] \gets |T|+1$
          \Comment{add new tips to tree and tip array}
          \State $\tip[u] \gets |T|+2$
          \Comment{corresponding to $u$ and $v$}
          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State delete $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees, via the
dot-product of the two trees' feature vectors in the space of all possible
subset trees with branch lengths (see \cref{subsec:treeshape}). {\color{blue}
\uline{Because the branch lengths are continuous, there are infinitely many
possible subset trees; hence, the feature space is infinite-dimensional.}}
The kernel was implemented using the fast algorithm developed by
\textcite{moschitti2006making}. First, the production rule of each node, which
is the total number of children and the number of leaf children, is recorded.
The nodes of both trees are ordered by production rule, and a list of pairs of
nodes sharing the same production rule is created. These are the nodes for
which the value of the tree kernel must be computed - all other pairs have a
value of zero. The pairs to be compared are then re-ordered so that the child
nodes are always evaluated before their parents. Due to its recursive
definition, ordering the pairs in this way allows the tree kernel to be
computed by dynamic programming. The complexity of this implementation is
$O(|T_1||T_2|)$ for the two trees $T_1$ and $T_2$ being compared.

The tree kernel cannot be used directly as a distance measure for \gls{ABC},
since it is maximized, not minimized, when the two trees being compared are the
same. Therefore, we defined the distance between two trees as
\[
  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
\]
which is a number between 0 and 1 that is minimized when $T_1 = T_2$. This is
similar to the normalization used by \textcite{collins2002new,
poon2013mapping}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of \textcite{del2012adaptive}, the
parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

We assumed that all contacts had symmetric transmission risk, which was
implemented by replacing each undirected edge in the network with two directed
edges (one in each direction). Nodes in our networks followed simple \gls{SI}
dynamics, meaning that they became infected at a rate proportional to their
number of infected neighbours, and never recovered. We did not consider the
time scale of the transmission trees in these simulations, only their shape.
Therefore, the transmission rate along each edge in the network was set to 1,
the removal rate of each node was set to 0, and all transmission trees' branch
lengths were scaled by their mean. The \textit{igraph} library's implementation
of the BA model~\autocite{csardi2006igraph} was used to generate the graphs.
The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
computer cluster. With the exception of our own programs, all analyses were
done in \software{R}, and all packages listed below are \software{R} packages.
{\color{blue}\uline{Code to run all simulation-based experiments is freely
available at \url{https://github.com/rmcclosk/thesis}.}}

@@

{\color{blue}\uline{
Our first computational experiment was designed as an exploratory analysis of
the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
any of the four parameters were identifiable from the shape of the transmission
tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
was varied one at a time, while holding the other parameters at fixed, known
values. Contact networks were generated according to each set of parameter
values, and transmission trees were simulated over the networks. We then
evaluated how well a classifier based on the tree kernel could differentiate
the trees simulated under distinct parameter values. If the classifier's
cross-validation accuracy was high, this could be taken as an indication that
the parameter in question was identifiable in the range of values considered. 
A caveat of this preliminary analysis is that, since all parameters but one
were held at known values, nothing could be said about the identifiability of
\emph{combinations} of parameters; this issue will be explored later by jointly
estimating all parameters with \gls{ABC}.}

\uline{In addition to testing for identifiability, a secondary objective of
this analysis was to validate the use of the tree kernel as a distance measure
for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
choice of distance function is extremely important for the accuracy of the
\gls{ABC} approximation to the posterior. Therefore, we evaluated two
additional tree statistics in the same manner as we evaluated the tree kernel
(that is, by constructing and testing a classifier). First, we considered
Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
index is widely used for characterizing
phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
between transmission trees simulated under different contact network
types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
lengths into account, considering only the tree's topology. The other statistic
we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
compares two trees based on normalized distributions of their branching times}
(see \cref{sec:treeshape}).  \uline{In contrast with Sackin's index, the
\gls{nltt} does not explicity consider the trees' topologies, but it does use
their normalized branch lengths. While the \gls{nltt} is a newly developed
statistic not yet in widespread use, the unnormalized
\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
extracting epidemiological information from
phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
classify the \gls{BA} parameters more accurately than either Sackin's index or
the \gls{nltt}, since the tree kernel takes both topology and branch lengths
into account. }}

This experiment involved a large number of variables that were varied
combinatorially. For ease of exposition, we will describe a single experiment
first, then enumerate the values of all variables for which the experiment was
repeated. The parameters of the tree kernel, $\lambda$ and $\sigma$
(\cref{subsec:treeshape}), will be referred to as \defn{meta-parameters} to
distinguish them from the parameters of the \gls{BA} model.

The attachment power parameter \gls{alpha} was varied among three values: 0.5,
1.0, and 1.5. For each value, the \software{sample\_pa} function in the
\software{igraph} package was used to simulate 100 networks, with the other
parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
300 networks. An epidemic was simulated on each network using our
\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
point 500 of them were sampled to form a transmission tree. A total of 300
transmission trees were thus obtained, comprised of 100 trees for each of the
three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
descending from the left child of each node was not smaller than that
descending from the right child. Summary statistics, such as Sackin's index and
the ratio of internal to terminal branch lengths, were computed for each
simulated tree using our \software{treestat} binary. The trees were visualized
using the \software{ape} package~\autocite{paradis2004ape}. {\color{blue}\uline{
Both the tree kernel and the \gls{nltt} are pairwise statistics, and the
\glspl{SVM} classifiers we used to investigate them operate on pairwise
distance matrices.}} Our \software{treekernel} binary was used to calculate the
value of the kernel for each pair of trees, with the meta-parameters set to
$\lambda = 0.3$ and $\sigma = 4$. These values were stored in a symmetric 300
$\times$ 300 kernel matrix.  Similarly, we computed the \gls{nltt} statistic
between each pair of trees using our \software{treestat} binary, and stored
them in a second $300 \times 300$ matrix.

@@

We simulated three trees each under a variety of parameter values and ran the
\software{netabc} program to estimate posterior distributions for the
parameters. The parameter values and priors used are listed in
\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\lambda = 0.3$
and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
sampled datasets per particle, and the $\alpha_{\ESS}$ parameter (not to be
confused with the \gls{BA} preferential attachment parameter, see
\cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
acceptance rate of the \gls{MH} kernel dropped below 1.5\%, the same criterion
used by \citeauthor{del2012adaptive}. Approximate marginal posterior densities
for each parameter were calculated using the \software{density} function in
\software{R} applied to the final weighted population of particles.
{\color{red}\sout{Credible intervals were obtained for each parameter using the
\software{HPDinterval} function in the \software{coda}
package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
obtained for each parameter using the \software{wtd.mean} function in the
\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
obtained using the \software{hpd} function in the \software{TeachingDemos}
package~\autocite{snow2013teachingdemos}.}}

@@

Two further experiments were performed to address potential sources of error.
To evaluate the effect of model misspecification in the case of heterogeneity
among nodes, we generated a network where half the nodes were attached with
power $\alpha$ = 0.5 and the other half with power $\alpha$ = 1.5. The other
parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
investigate the effects of potential sampling
bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
the tips were sampled in a peer-driven fashion, rather than at random. That is,
the probability to sample a node was twice as high if any of that node's
network peers had already been sampled. The parameters of this network were $N$
= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.

@@

Trees simulated under different values of \gls{alpha} were visibly quite
distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes, which, once
infected, are likely to transmit to many other nodes. This results in a more
unbalanced, ladder-like structure in the phylogeny, compared to networks with
lower \gls{alpha} values. None of the other three parameters produced trees
that were as easily distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).  Sackin's index, which
measures tree imbalance, was significantly correlated with all four parameters
    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
     \Sexpr{round(I.sackin.cor$estimate, 2)},
     \Sexpr{round(m.sackin.cor$estimate, 2)},
     \Sexpr{round(N.sackin.cor$estimate, 2)};
     $p$-values
     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
     $\Sexpr{pp(I.sackin.cor$p.value)}$,
     $\Sexpr{pp(m.sackin.cor$p.value)}$,
     $\Sexpr{pp(N.sackin.cor$p.value)}$).
The ratio of internal to terminal branch lengths was negatively correlated with
\gls{alpha} and \gls{I}, and positively correlated with \gls{m} and \gls{N}
  (Spearman's rho
    \Sexpr{round(alpha.ratio.cor$estimate, 2)},
    \Sexpr{round(I.ratio.cor$estimate, 2)},
    \Sexpr{round(m.ratio.cor$estimate, 2)},
    \Sexpr{round(N.ratio.cor$estimate, 2)};
  all $p < 10^{-5}$).

@@

\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
first two principal components of the kernel matrix. The figure shows only the
simulations with 500-tip trees and 1000 infected nodes. The three \gls{alpha}
and \gls{I} values considered are well separated from each other in feature
space. On the other hand, the three \gls{N} values overlap significantly, and
the three \gls{m} values are virtually indistinguishable. Similar observations
can be made for other values of \gls{I} and the number of tips
(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
\gls{N} separated more clearly with larger numbers of tips, and in the case of
\gls{N}, with larger epidemic sizes {\color{blue}(\cref{fig:Ikpca,fig:Nkpca})}.

@@

The accuracy of classifiers for $I$ varied significantly with the number of
tips in the tree. For 100-tip trees, the average $R^2$ values were
  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
trees, the values increased to
  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
Finally, the performance of classifiers for $N$ depended heavily on the
epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
for the smallest epidemic and smallest sample size, to
  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
to
  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
Sackin's index did not accurately classify $N$ in any scenario, with an average
$R^2$ of
  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
and little variation between scenarios.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
  \caption[Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and
  Sackin's index regression classifiers for BA model parameters.]{
      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
      using \gls{nltt} (centre), and linear regression using Sackin's index
      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
      simulated transmission trees over networks with three different values of
      the parameter being tested, assuming perfect knowledge of the other
      parameters. Vertical lines are empirical 95\% confidence intervals based
      on 1000 two-fold cross-validations. {\color{blue}\uline{ The classifiers
      for \gls{I} were not evaluated with 1000-tip trees, because one of the
      tested \gls{I} values was 500, and it is not possible to sample a tree of
      size 1000 from 500 infected individuals.}}
  }
  \label{fig:rsquared}
\end{figure}
